### 2020-01-30 wuhan.tex
Have to call my Mom

\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\new}[1]{\textcolor{blue}{#1}}

@@

Since December 2019, the novel coronavirus (2019-nCoV) has been
spreading in China and other parts of the world.
As of January 28th, 2020, the World Health Organization has
confirmed 4593 cases, including 56 confirmed cases in 14 different
countries, outside China \citep{who28report}.
Although the virus is believed to have originated
from animal reservoirs, a recent case from Viet Nam
demonstrated its ability to directly transmit between
humans \citep{who26report},
posing a greater threat for its spread.

As the disease continues to spread, 
many researchers are rushing to publish their 
analysis of the outbreak and, in particular, their
estimates of the basic reproductive number $\mathcal R_0$ (i.e., the 
average number of secondary cases generated 
by a primary case in a fully susceptible population \citep{anderson1991infectious, diekmann1990definition}).
Estimating the basic reproductive number is of interest during an outbreak
because it allows prediction about the final size of the outbreak \citep{anderson1991infectious, ma2006generality, arino2007final, andreasen2011final, miller2012note} and the required level of intervention \citep{anderson1991infectious}.
However, it can be difficult to assess the validity of the estimates of $\mathcal R_0$ (as well as the associated degrees of uncertainty)
when the estimation methods and their underlying assumptions vary widely,
especially since these assumptions can directly affect the estimates.

@@

Early in an outbreak, $\mathcal R_0$ cannot be estimated directly;
instead, $\mathcal R_0$ is often inferred from
the exponential growth rate $r$, which can be estimated reliably from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
Given an estimate of the exponential growth rate $r$ and a distribution $g(\tau)$ of
generation intervals (i.e., the time between when a person become 
infected and that person infects another person \citep{svensson2007note}), the basic reproductive
number can be estimated via the Euler-Lotka equation \citep{wallinga2007generation}:
\begin{equation}
1/\mathcal R_0 = \int \exp(-r\tau) g(\tau) d\tau.
\end{equation}
In other words, estimates of $\mathcal R_0$ must
depend on the assumptions about the
exponential growth rate $r$ and the shape of the generation-interval distribution $g(\tau)$.

Here, we use the gamma approximation framework \citep{park2019practical} to (1) 
characterize the
amount of uncertainty present in the exponential growth rates and the shape of the generation-interval distribution
and (2) assess the 
degree to which these uncertainties
affect the estimate of $\mathcal R_0$.
Assuming that generation intervals follow a gamma distribution
with the mean $\bar G$ and the squared coefficient of variation (CV$^2$) $\kappa$, 
we have
\begin{equation}
\mathcal R_0 = \left(1 + \kappa r \bar{G}\right)^{1/\kappa}.
\label{eq:gamma}
\end{equation}
This equation demonstrates that a generation-interval distribution
that has a larger mean (higher $\bar{G}$) or is less variable (lower $\kappa$)
will give a higher estimate of $\mathcal R_0$ for the same value of $r$.

@@

We calculate the pooled estimates for each parameter ($r$, $\bar G$, and $\kappa$) with a Bayesian multilevel modeling approach, which assumes that the parameters across different studies come from the same Gamma distribution:
\begin{equation}
\begin{aligned}
r_i &\sim \mathrm{Gamma}(\mathrm{mean}=\mu_r, \mathrm{shape}=\alpha_r),\\
\bar{G}_i &\sim \mathrm{Gamma}(\mathrm{mean}=\mu_G, \mathrm{shape}=\alpha_G),\\
\kappa_i &\sim \mathrm{Exponential}(\mathrm{mean}=\mu_\kappa),\\
\end{aligned}
\end{equation}
We account for uncertainties associated with $r_i$, $\bar G_i$ and $\kappa_i$ (and their correlations), by drawing random samples at each Metropolis-Hastings step from 100,000 samples that we calculated earlier;
this approach is analogous to Bayesian methods for analyzing phylogenetic data, which pften rely on drawing random samples of phylogenetic trees from a discrete set to account for phylogenetic uncertainty \citep{pagel2004bayesian,bedford2014integrating}.
Weak priors are assumed on the hyperparameters:
\begin{equation}
(\mu_r, \alpha_r, \mu_G, \alpha_G, \mu_\kappa) \sim \mathrm{Gamma}(\mathrm{shape}=0.1, \mathrm{rate}=0.1).
\end{equation}
We run 4 parallel Markov Chain Monte Carlo (MCMC) chains that consist of 200,000 burnin steps and 200,000 sampling steps.
Posterior samples are thinned every 400 steps.
Convergence is assessed by ensuring that the Gelman-Rubin statistic is below 1.01 for all parameters \citep{gelman1992inference}.

>>>>

\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\jd}[1]{\comment{magenta}{JD}{#1}}
\newcommand{\new}[1]{\textcolor{blue}{#1}}

@@

Since December 2019, the novel coronavirus (2019-nCoV) has been
spreading in China and other parts of the world.
As of January 28th, 2020, the World Health Organization has
confirmed 4593 cases, including 56 confirmed cases in 14 different
countries, outside China \citep{who28report}.
Although the virus is believed to have originated
from animal reservoirs, a recent case from Viet Nam
demonstrated its ability to directly transmit between
humans \citep{who26report},
posing a greater threat for its spread.
\jd{It is spread primarily from person to person; that's been obvious for a while and is now confirmed. We should find a better cite and leave Viet Nam out of it.}

\jd{What's up with the line breaks??}
As the disease continues to spread, 
many researchers have published
analyses of the outbreak, many focusing in particular on
estimates of the basic reproductive number $\mathcal R_0$ (i.e., the 
average number of secondary cases generated 
by a primary case in a fully susceptible population \citep{anderson1991infectious, diekmann1990definition}).
Estimating the basic reproductive number is of interest during an outbreak
because it provides information about the level of intervention required to interrupt transmission \citep{anderson1991infectious}, and about the final size of the outbreak \citep{anderson1991infectious, ma2006generality}.
\jd{We really only need one cite for this whole s., but I think Ma will make David happy.}
However, it can be difficult to assess the validity of the estimates of $\mathcal R_0$ (as well as the associated degrees of uncertainty)
when the estimation methods and their underlying assumptions vary widely,
especially since these assumptions can directly affect the estimates.

@@

Early in an outbreak, $\mathcal R_0$ cannot be estimated directly;
instead, $\mathcal R_0$ is often inferred from
the exponential growth rate $r$, which can be estimated reliably from incidence data \citep{mills2004transmissibility, nishiura2009transmission, ma2014estimating}. \jd{Trimmed references; in particular, don't really want to point people to Chowell for stats.}
Given an estimate of the exponential growth rate $r$ and a distribution $g(\tau)$ of
generation intervals (i.e., the time between when a person become 
infected and that person infects another person \citep{svensson2007note}), the basic reproductive
number can be estimated via the Euler-Lotka equation \citep{wallinga2007generation}:
\begin{equation}
1/\mathcal R_0 = \int \exp(-r\tau) g(\tau) d\tau.
\end{equation}
In other words, estimates of $\mathcal R_0$ must
depend on the assumptions about the
exponential growth rate $r$ and the shape of the generation-interval distribution $g(\tau)$.

Here, we use the gamma approximation framework \citep{park2019practical} to (1) 
characterize the
amount of uncertainty present in the exponential growth rates and the shape of the generation-interval distribution
and (2) assess the 
degree to which these uncertainties
affect the estimate of $\mathcal R_0$.
\jd{We should cite another group as well about this approximation.}
Assuming that generation intervals follow a gamma distribution
with the mean $\bar G$ and the squared coefficient of variation (CV$^2$) $\kappa$, 
we have
\begin{equation}
\mathcal R_0 = \left(1 + \kappa r \bar{G}\right)^{1/\kappa}.
\label{eq:gamma}
\end{equation}
This equation demonstrates that a generation-interval distribution
that has a larger mean (higher $\bar{G}$) or is less variable (lower $\kappa$)
will give a higher estimate of $\mathcal R_0$ for the same value of $r$.

@@

\jd{Are you still using exponential?}
We construct pooled estimates for each parameter ($r$, $\bar G$, and $\kappa$) using a Bayesian multilevel modeling approach, which assumes that the parameters across different studies come from the same Gamma distribution:
\begin{equation}
\begin{aligned}
r_i &\sim \mathrm{Gamma}(\mathrm{mean}=\mu_r, \mathrm{shape}=\alpha_r),\\
\bar{G}_i &\sim \mathrm{Gamma}(\mathrm{mean}=\mu_G, \mathrm{shape}=\alpha_G),\\
\kappa_i &\sim \mathrm{Exponential}(\mathrm{mean}=\mu_\kappa),\\
\end{aligned}
\end{equation}
We account for uncertainties associated with $r_i$, $\bar G_i$ and $\kappa_i$ (and their correlations), by drawing random samples at each Metropolis-Hastings step from 100,000 samples that we calculated earlier;
this approach is analogous to Bayesian methods for analyzing phylogenetic data, which pften rely on drawing random samples of phylogenetic trees from a discrete set to account for phylogenetic uncertainty \citep{pagel2004bayesian,bedford2014integrating}.
\jd{use mean (not rate) and shape; specify units for the means.}
Weak priors are assumed on the hyperparameters:
\begin{equation}
(\mu_r, \alpha_r, \mu_G, \alpha_G, \mu_\kappa) \sim \mathrm{Gamma}(\mathrm{shape}=0.1, \mathrm{rate}=0.1).
\end{equation}
We run 4 parallel Markov Chain Monte Carlo (MCMC) chains that consist of 200,000 burnin steps and 200,000 sampling steps.
Posterior samples are thinned every 400 steps.
Convergence is assessed by ensuring that the Gelman-Rubin statistic is below 1.01 for all parameters \citep{gelman1992inference}.
